Unprepared colonoscopy identifies microbiota specific to the proximal and distal human colon

### 2017-07-27 submission/manuscript.tex
code tweaks to make it run

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
versus distal lumen (AUC = 0.8). OTUs that were differentially abundant
between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis} (XX\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors.
((14)). The \emph{Fusobacterium} positive samples were located in x\% of
the the proximal and X\% of the distal mucosa and represented as much as
1\% of any sample (Figure 6A). OTU152 was similar to the members of the
\emph{Porphyromonas} genus and the most frequent sequence in that OTU
aligned to \emph{Porphyromonas asacharolytica} (X\% over full length),
another bacterium commonly detected and isolated from colorectal tumors.
OTU152 was only detected on the distal mucosa, and in fact was one of
the OTUs the classification model identified as separating distal and
proximal sides (Figure 6B). Among the samples that were positive for
\emph{P. asacharolytica}, the relative abundances for this OTU ranged
from 0.01\% - 16\%. Thus, disease-associated OTUs could be found in our
sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each
site, there was a wide amount of variation for each phyla with
communities primarily dominated by the \emph{Bacteriodes} and
\emph{Firmicutes}. This likely reflects not only the variability between
human subjects, casued by differences in age, gender, diet, but also
reports of biogeogrphical ``patchiness'' in the gut microbiome. Several
studies have noted that the bacteria recoverable from the same mucosal
sample location can be vastly different when the samples are taken just
1 cm away from each other (15). Similar patchiness is also observed in
lumenal contents and fecal samples themselves; there is observed
separation of different interacting microbes along the length of a stool
sample, for instance (16). That said, across our samples the mucosal
samples harbor more \emph{Proteobacteria}, consistent with previous
studies comparing mucosal swabs to lumenal content in humans (5). Hence,
the conclusions we can draw from phyla analysis are likely impacted by
patchiness between subjects.

To get around the noisiness from a diverse set of samples, we built a
Random Forest model to identify microbes specific to each side. For each
comparison we identified top X OTUs that were strongly predictive of one
site or another. Generally, OTUs identified in each location were
consistent with known physiological gradients along the gut axis (6).
For instance, the proximal mucosa contains the highest oxygen
concentratons of the colon and harbored mucosa-associated facultative
anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
host strictly anaerobic species such as \emph{Porphyromonas},
\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. The
model was less effective at classifying the proximal and distal lumenal
contents, probably because the samples are arguably composed of the same
bacteria but differ in water content.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (8) and have oncogenic
properties in cell culture and in mice (17). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous distal mucosa ((5)). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreases along the colon to rectum ((18)). Of the 8 (40\%) individuals
positive for \emph{F. nucleatum} in this study, the bacterium was spread
across the proximal mucosa, distal lumen and distal mucosa. The presence
of \emph{F. nucleatum} in a healthy individual is not necessarily linked
to the development of future colorectal cancers (cite). Data examining
bacterial biofilms on CRC tumors suggests that \emph{Fusobacteria}
species are more commonly found on proximal tumors and in biofilms,
indicating that it is not only the presence of the bacteria but the
organization of the tumor community that contributes to
\emph{Fusobacterium's} role in tumorigenesis ((7)). Finally,
\emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
not only co-occur on CRC tumors but also to synergistically promote
tumorigenesis in an oral cancer model ((19), (20)). Thus, further
analysis of the distribution and activities of these pathogens may
elucidate a mechanism for development of IBD or CRC subtypes in the
proximal or distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD
(cite). Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-a production than strains from healthy individuals ((21)),
suggesting the bacteria may increase inflammation in the gut as well
(cite - Dharmani 2011). \emph{F. varium} isolated from UC patients
caused colonic ulcers in an experimental mouse model (Ohkusa 2003).
\emph{F. varium} was only detected in 3 of our study participants and
two of those samples were isolated form the proximal mucosa (cite
figure). Although there is low occurrance in our study, \emph{F. varium}
is most commonly isolated from UC patient biopsies from the ileum or
cecum (cite), suggesting this species may exhibit preference for the
different environmental conditions of these gastrointestinal sites.
Further work will assess how gut environment may select for species
which then cause localized disease.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing ((22)). This group
found that mucosal and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site ((22)). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. Bowel preparation clearly
induces bias into the microbes recovered from sampling the lumen or
mucosa of a prepared bowel. Thus our findings in this study are
strengthened by the lack of bowel preparation.

@@

At a baseline visit, subjects were consented and given a home collection
stool kit (Source of kit supplies). At least one week prior to the
scheduled colonoscopy, subjects were to collect whole stool at home and
ship the samples to the University on ice. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On the procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Patients were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and endoscopy
brush used to collect lumenal/stool contents. Two lumenal samples were
collected and the contents immediately deposited into RNAlater (source)
and flash-frozen in liquid nitrogen. The brushes were withdrawn and
biopsy forceps were used to collect mucosal biopsies on sections of the
colon that were pink and free of stool matter. Three mucosal biopsies
were collected and flash-frozen in RNAlater. These samples comprised the
distal or distal colon samples. The sigmoidoscope was then withdrawn and
a pediatric colonoscope was inserted to reach the ascending colon.
Samples were then collected as in the distal colon and the colonoscope
withdrawn. All samples were stored at -80 C until study completion.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described ((23)). Sequences were curated
using the mothur software as described previously ((24)). The sequences
were assigned taxonomic classification using a naive Bayesian classifier
trained using a 16S rRNA gene training set from the Ribosomal Database
Project (RDP) ((25)) and clustered into operational taxonomic units
(OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
mock community revealed the error rate to be X\%. Samples were rarefied
to 4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator
and theta YC calculator metrics in mothur version
\textless{}3.2?\textgreater{} ((24)). ThetaYC distances were calculated
to determine the dissimilarity between two samples. Random Forest
classification models were built using the randomForest R package and
resultant models were used to identify the OTUs that were most important
for classifying each location ((26)). To get species-level information
about sequences of interest, sequences were aligned using blastn and the
species name was only used if the identity score was \textgreater{}=
99\%.

@@

Differences in community membership at the phyla level were tested using
the analysis of molecular variance (AMOVA) metric in mothur. Differences
in thetaYC distances by location were tested using the Wilcoxon rank-sum
test adjusted for multiple comparisons using the Benjamini-Hochberg
procedure.

\paragraph{Data availability}\label{data-availability}

\paragraph{Github repository}\label{github-repository}

@@

Location and abundance of cancer-associated OTUs. Relative abundance was
calculated and plotted by sample site for each OTU of interest: (A)
\emph{Fusobacterium nucleatum} and (B) \emph{Porphyromonas
asacharolytica}

@@

\hypertarget{ref-Hong2011}{}
15. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
\textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
analysis of the mucosal microbiota in healthy individuals reveals
ubiquitous bacterial groups and micro-heterogeneity. PLoS ONE
\textbf{6}:e25042.
doi:\href{https://doi.org/10.1371/journal.pone.0025042}{10.1371/journal.pone.0025042}.

\hypertarget{ref-Stearns2011}{}
16. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
\textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
\textbf{Croitoru K}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}.
2011. Bacterial biogeography of the human digestive tract. Scientific
Reports \textbf{1}.
doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.

\hypertarget{ref-Sears2014}{}
17. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
doi:\href{https://doi.org/10.1016/j.chom.2014.02.007}{10.1016/j.chom.2014.02.007}.

\hypertarget{ref-Mima2016}{}
18. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
\textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
\textbf{Silva A da}, \textbf{Gu M}, \textbf{Li W}, \textbf{Hamada T},
\textbf{Kosumi K}, \textbf{Hanyuda A}, \textbf{Liu L}, \textbf{Kostic
AD}, \textbf{Giannakis M}, \textbf{Bullman S}, \textbf{Brennan CA},
\textbf{Milner DA}, \textbf{Baba H}, \textbf{Garraway LA},
\textbf{Meyerhardt JA}, \textbf{Garrett WS}, \textbf{Huttenhower C},
\textbf{Meyerson M}, \textbf{Giovannucci EL}, \textbf{Fuchs CS},
\textbf{Nishihara R}, \textbf{Ogino S}. 2016. Fusobacterium nucleatum in
colorectal carcinoma tissue according to tumor location. Clinical and
Translational Gastroenterology \textbf{7}:e200.
doi:\href{https://doi.org/10.1038/ctg.2016.53}{10.1038/ctg.2016.53}.

\hypertarget{ref-Whitmore2014}{}
19. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
cancer. PLoS Pathogens \textbf{10}:e1003933.
doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.

\hypertarget{ref-Flynn2016}{}
20. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
Metabolic and community synergy of oral bacteria in colorectal cancer.
mSphere \textbf{1}:e00102--16.
doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.

\hypertarget{ref-Strauss2011}{}
21. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
gut mucosa-derived fusobacterium nucleatum positively correlates with
IBD status of the host. Inflammatory Bowel Diseases
\textbf{17}:1971--1978.
doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.

\hypertarget{ref-Shobar2016}{}
22. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
\textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
\textbf{Losurdo J}, \textbf{Engen PA}, \textbf{Sun Y}, \textbf{Koenig
L}, \textbf{Mutlu EA}. 2016. The effects of bowel preparation on
microbiota-related metrics differ in health and in inflammatory bowel
disease and for the mucosal and luminal microbiota compartments.
Clinical and Translational Gastroenterology \textbf{7}:e143.
doi:\href{https://doi.org/10.1038/ctg.2015.54}{10.1038/ctg.2015.54}.

\hypertarget{ref-Kozich2013}{}
23. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Schloss2009}{}
24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Wang2007}{}
25. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
of rRNA sequences into the new bacterial taxonomy. Applied and
Environmental Microbiology \textbf{73}:5261--5267.
doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.

\hypertarget{ref-Liaw2002}{}
26. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News: The Newsletter of the R Project
\textbf{2}:18--22.

>>>>

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.9) compared to the proximal
versus distal lumen (AUC = 0.8). The model that separated proximal and
distal mucosal samples used 6 features. OTUs that were differentially
abundant between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B). The
model used 44 features to separate proximal and distal lumenal samples.
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors.
((14)). There were 14 samples with the \emph{F. nucleatum subsp.
animalis} sequence. Four sequences were located in the proximal mucosa
and three in the distal mucosal samples (Figure 6A). The second most
frequent \emph{Fusobacterium} sequence was OTU472, which aligns with
99\% identity to \emph{F. varium}. In addition to \emph{F. nucleatum},
\emph{F. varium} has been associated with IBD ((15)). 4 study
participants harbored \emph{F. varium} and the samples were split evenly
between the proximal and distal mucosa (Figure 6B). OTU152 was similar
to the members of the \emph{Porphyromonas} genus and the most frequent
sequence in that OTU aligned to \emph{Porphyromonas asacharolytica}
(99\% over full length), another bacterium commonly detected and
isolated from colorectal tumors. OTU152 was only detected on the distal
mucosa, and in fact was one of the OTUs the classification model
identified as separating distal and proximal sides (Figure 6C). Among
the samples that were positive for \emph{P. asacharolytica}, the
relative abundances for this OTU ranged from 0.01\% - 16\%. Thus,
disease-associated OTUs could be found in our sample set of 20 healthy
individuals.

@@

When examining the relative abundance of the different phyla at each
site, there was a wide amount of variation for each phyla with
communities primarily dominated by the \emph{Bacteriodes} and
\emph{Firmicutes}. This likely reflects not only the variability between
human subjects, casued by differences in age, gender, diet, but also
reports of biogeogrphical ``patchiness'' in the gut microbiome. Several
studies have noted that the bacteria recoverable from the same mucosal
sample location can be vastly different when the samples are taken just
1 cm away from each other (16). Similar patchiness is also observed in
lumenal contents and fecal samples themselves; there is observed
separation of different interacting microbes along the length of a stool
sample, for instance (17). That said, across our samples the mucosal
samples harbor more \emph{Proteobacteria}, consistent with previous
studies comparing mucosal swabs to lumenal content in humans (5). Hence,
the conclusions we can draw from phyla analysis are likely impacted by
patchiness between subjects.

To get around the noisiness from a diverse set of samples, we built a
Random Forest model to identify microbes specific to each side. For each
comparison we identified top 5 OTUs that were strongly predictive of one
site or another. Generally, OTUs identified in each location were
consistent with known physiological gradients along the gut axis (6).
For instance, the proximal mucosa contains the highest oxygen
concentratons of the colon and harbored mucosa-associated facultative
anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
host strictly anaerobic species such as \emph{Porphyromonas},
\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. The
model was less effective at classifying the proximal and distal lumenal
contents, probably because the samples are arguably composed of the same
bacteria but differ in water content.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (8) and have oncogenic
properties in cell culture and in mice (18). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous distal mucosa (5). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreases along the colon to rectum (19). Of the 8 (40\%) individuals
positive for \emph{F. nucleatum} in this study, the bacterium was spread
across the proximal mucosa, distal lumen and distal mucosa. Data
examining bacterial biofilms on the mucosa of CRC tumors suggests that
\emph{Fusobacteria} species are more commonly found on proximal tumors
and in biofilms, indicating that it is not only the presence of the
bacteria but the organization of the tumor community that contributes to
\emph{Fusobacterium's} role in tumorigenesis (7). Finally,
\emph{Fusobacterium} and \emph{Porphyromonas} species have been known to
not only co-occur on CRC tumors but also to synergistically promote
tumorigenesis in an oral cancer model (20), (21). Thus, further analysis
of the distribution and activities of these pathogens may elucidate a
mechanism for development of IBD or CRC subtypes in the proximal or
distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (15).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-a production than strains from healthy individuals (22),
suggesting the bacteria may increase inflammation in the gut as well
(23). \emph{F. varium} isolated from UC patients caused colonic ulcers
in an experimental mouse model (24). \emph{F. varium} was only detected
in 3 of our study participants and two of those samples were isolated
form the proximal mucosa (Figure 6B). Although there is low occurrance
in our study, \emph{F. varium} is most commonly isolated from UC patient
biopsies from the ileum or cecum (25), suggesting this species may
exhibit preference for the different environmental conditions of these
gastrointestinal sites. Further work will assess how gut environment may
select for species which then cause localized disease.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing (26). This group
found that mucosal and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site (26). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. Bowel preparation clearly
induces bias into the microbes recovered from sampling the lumen or
mucosa of a prepared bowel. Thus our findings in this study are
strengthened by the lack of bowel preparation.

@@

At a baseline visit, subjects were consented and given a home collection
stool kit (Source of kit?). At least one week prior to the scheduled
colonoscopy, subjects were to collect whole stool at home and ship the
samples to the University on ice. Notably, subjects did not undergo any
bowel preparation method prior to sampling. On the procedure day,
subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Patients were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and endoscopy
brush used to collect lumenal/stool contents. Two lumenal samples were
collected and the contents immediately deposited into RNAlater (Fischer)
and flash-frozen in liquid nitrogen. The brushes were withdrawn and
biopsy forceps were used to collect mucosal biopsies on sections of the
colon that were pink and free of stool matter. Three mucosal biopsies
were collected and flash-frozen in RNAlater. These samples comprised the
distal or distal colon samples. The sigmoidoscope was then withdrawn and
a pediatric colonoscope was inserted to reach the ascending colon.
Samples were then collected as in the distal colon and the colonoscope
withdrawn. All samples were stored at -80 C until study completion.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (27). Sequences were curated
using the mothur software as described previously (28). The sequences
were assigned taxonomic classification using a naive Bayesian classifier
trained using a 16S rRNA gene training set from the Ribosomal Database
Project (RDP) (29) and clustered into operational taxonomic units (OTUs)
based on a 97\% similarity cutoff. Sequencing and analysis of a mock
community revealed the error rate to be 1.8\%. Samples were rarefied to
4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator
and theta YC calculator metrics in mothur version 3.2? (28). ThetaYC
distances were calculated to determine the dissimilarity between two
samples. Random Forest classification models were built using the
randomForest R package and resultant models were used to identify the
OTUs that were most important for classifying each location (30). To get
species-level information about sequences of interest, sequences were
aligned using blastn and the species name was only used if the identity
score was \textgreater{}= 99\%.

@@

Differences in community membership at the phyla level were tested using
the analysis of molecular variance (AMOVA) metric in mothur. Differences
in thetaYC distances by location were tested using the Wilcoxon rank-sum
test adjusted for multiple comparisons using the Benjamini-Hochberg
procedure. \#\#\#\# Data availability 16S rRNA gene sequence reads and
experiment metadata are available on the NCBI Sequence Read Archive
(SRA) with accession number XXXX. A reproducible data analysis pipeline
can be found at
\url{https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017}.

@@

Location and abundance of cancer-associated OTUs. Relative abundance was
calculated and plotted by sample site for each OTU of interest: (A)
\emph{Fusobacterium nucleatum subsp. animalis} (B) \emph{Fusobacterium
varium} and (C) \emph{Porphyromonas asacharolytica}

@@

\hypertarget{ref-Lee2016}{}
15. \textbf{Lee Y}, \textbf{Eun CS}, \textbf{Lee AR}, \textbf{Park CH},
\textbf{Han DS}. 2016. FusobacteriumIsolates recovered from colonic
biopsies of inflammatory bowel disease patients in korea. Annals of
Laboratory Medicine \textbf{36}:387.
doi:\href{https://doi.org/10.3343/alm.2016.36.4.387}{10.3343/alm.2016.36.4.387}.

\hypertarget{ref-Hong2011}{}
16. \textbf{Hong P-Y}, \textbf{Croix JA}, \textbf{Greenberg E},
\textbf{Gaskins HR}, \textbf{Mackie RI}. 2011. Pyrosequencing-based
analysis of the mucosal microbiota in healthy individuals reveals
ubiquitous bacterial groups and micro-heterogeneity. PLoS ONE
\textbf{6}:e25042.
doi:\href{https://doi.org/10.1371/journal.pone.0025042}{10.1371/journal.pone.0025042}.

\hypertarget{ref-Stearns2011}{}
17. \textbf{Stearns JC}, \textbf{Lynch MDJ}, \textbf{Senadheera DB},
\textbf{Tenenbaum HC}, \textbf{Goldberg MB}, \textbf{Cvitkovitch DG},
\textbf{Croitoru K}, \textbf{Moreno-Hagelsieb G}, \textbf{Neufeld JD}.
2011. Bacterial biogeography of the human digestive tract. Scientific
Reports \textbf{1}.
doi:\href{https://doi.org/10.1038/srep00170}{10.1038/srep00170}.

\hypertarget{ref-Sears2014}{}
18. \textbf{Sears CL}, \textbf{Garrett WS}. 2014. Microbes, microbiota,
and colon cancer. Cell Host \& Microbe \textbf{15}:317--328.
doi:\href{https://doi.org/10.1016/j.chom.2014.02.007}{10.1016/j.chom.2014.02.007}.

\hypertarget{ref-Mima2016}{}
19. \textbf{Mima K}, \textbf{Cao Y}, \textbf{Chan AT}, \textbf{Qian ZR},
\textbf{Nowak JA}, \textbf{Masugi Y}, \textbf{Shi Y}, \textbf{Song M},
\textbf{Silva A da}, \textbf{Gu M}, \textbf{Li W}, \textbf{Hamada T},
\textbf{Kosumi K}, \textbf{Hanyuda A}, \textbf{Liu L}, \textbf{Kostic
AD}, \textbf{Giannakis M}, \textbf{Bullman S}, \textbf{Brennan CA},
\textbf{Milner DA}, \textbf{Baba H}, \textbf{Garraway LA},
\textbf{Meyerhardt JA}, \textbf{Garrett WS}, \textbf{Huttenhower C},
\textbf{Meyerson M}, \textbf{Giovannucci EL}, \textbf{Fuchs CS},
\textbf{Nishihara R}, \textbf{Ogino S}. 2016. Fusobacterium nucleatum in
colorectal carcinoma tissue according to tumor location. Clinical and
Translational Gastroenterology \textbf{7}:e200.
doi:\href{https://doi.org/10.1038/ctg.2016.53}{10.1038/ctg.2016.53}.

\hypertarget{ref-Whitmore2014}{}
20. \textbf{Whitmore SE}, \textbf{Lamont RJ}. 2014. Oral bacteria and
cancer. PLoS Pathogens \textbf{10}:e1003933.
doi:\href{https://doi.org/10.1371/journal.ppat.1003933}{10.1371/journal.ppat.1003933}.

\hypertarget{ref-Flynn2016}{}
21. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
Metabolic and community synergy of oral bacteria in colorectal cancer.
mSphere \textbf{1}:e00102--16.
doi:\href{https://doi.org/10.1128/msphere.00102-16}{10.1128/msphere.00102-16}.

\hypertarget{ref-Strauss2011}{}
22. \textbf{Strauss J}, \textbf{Kaplan GG}, \textbf{Beck PL},
\textbf{Rioux K}, \textbf{Panaccione R}, \textbf{DeVinney R},
\textbf{Lynch T}, \textbf{Allen-Vercoe E}. 2011. Invasive potential of
gut mucosa-derived fusobacterium nucleatum positively correlates with
IBD status of the host. Inflammatory Bowel Diseases
\textbf{17}:1971--1978.
doi:\href{https://doi.org/10.1002/ibd.21606}{10.1002/ibd.21606}.

\hypertarget{ref-Dharmani2011}{}
23. \textbf{Dharmani P}, \textbf{Strauss J}, \textbf{Ambrose C},
\textbf{Allen-Vercoe E}, \textbf{Chadee K}. 2011. Fusobacterium
nucleatum infection of colonic cells stimulates MUC2 mucin and tumor
necrosis factor alpha. Infection and Immunity \textbf{79}:2597--2607.
doi:\href{https://doi.org/10.1128/iai.05118-11}{10.1128/iai.05118-11}.

\hypertarget{ref-Ohkusa2003}{}
24. \textbf{Ohkusa T}. 2003. Induction of experimental ulcerative
colitis by fusobacterium varium isolated from colonic mucosa of patients
with ulcerative colitis. Gut \textbf{52}:79--83.
doi:\href{https://doi.org/10.1136/gut.52.1.79}{10.1136/gut.52.1.79}.

\hypertarget{ref-Ohkusa2002}{}
25. \textbf{Ohkusa T}, \textbf{Sato N}, \textbf{Ogihara T},
\textbf{Morita K}, \textbf{Ogawa M}, \textbf{Okayasu I}. 2002.
Fusobacterium varium localized in the colonic mucosa of patients with
ulcerative colitis stimulates species-specific antibody. Journal of
Gastroenterology and Hepatology \textbf{17}:849--853.
doi:\href{https://doi.org/10.1046/j.1440-1746.2002.02834.x}{10.1046/j.1440-1746.2002.02834.x}.

\hypertarget{ref-Shobar2016}{}
26. \textbf{Shobar RM}, \textbf{Velineni S}, \textbf{Keshavarzian A},
\textbf{Swanson G}, \textbf{DeMeo MT}, \textbf{Melson JE},
\textbf{Losurdo J}, \textbf{Engen PA}, \textbf{Sun Y}, \textbf{Koenig
L}, \textbf{Mutlu EA}. 2016. The effects of bowel preparation on
microbiota-related metrics differ in health and in inflammatory bowel
disease and for the mucosal and luminal microbiota compartments.
Clinical and Translational Gastroenterology \textbf{7}:e143.
doi:\href{https://doi.org/10.1038/ctg.2015.54}{10.1038/ctg.2015.54}.

\hypertarget{ref-Kozich2013}{}
27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Schloss2009}{}
28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Wang2007}{}
29. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
of rRNA sequences into the new bacterial taxonomy. Applied and
Environmental Microbiology \textbf{73}:5261--5267.
doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.

\hypertarget{ref-Liaw2002}{}
30. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News: The Newsletter of the R Project
\textbf{2}:18--22.
